High-order cognition is supported by information-rich but compressible brain activity patterns

### 2023-03-02 paper/main.tex
applied a coat of polish to the text-- fixed a bunch of minor issues throughout (mostly typos).  i also corrected some errors in figure 1.

\title{High-order cognition is supported by complex but compressible brain activity patterns}

@@

We applied dimensionality reduction algorithms and pattern classifiers to
functional neuroimaging data collected as participants listened to a story,
temporally scrambled versions of the story, or underwent a resting state
scanning session. These experimental conditions were intended to require
different depths of processing and inspire different levels of cognitive
engagement. We considered two primary aspects of the data. First, we treated
the number of features (components) required to achieve a threshold decoding
accuracy as a proxy for the ``compressibility'' of the neural patterns (where
fewer components indicate greater compression). Second, we treated the
maximum achievable decoding accuracy across participants as an indicator of the
``stability'' of the recorded patterns. Overall, we found that neural patterns
recorded as participants listened to the intact story required fewer features
to achieve comparable classification accuracy to the other experimental
conditions. However, the peak decoding accuracy (achievable with more features)
was also highest during intact story listening. Taken together, our work
suggests that our brain networks flexibly reconfigure according to ongoing task
demands, and that the activity patterns associated with higher-order cognition
and high engagement are both more complex and more compressible than the
activity patterns associated with lower-order tasks and lower levels of
engagement.

@@

Large-scale networks, including the human brain, may be conceptualized as
occupying one or more positions along on a continuum. At one extreme, every
node is fully independent of every other node. At the other extreme, all nodes
behave identically. Each extreme optimizes key properties of how the network
functions. When every node is independent, the network is maximally
\textit{expressive}: if we define the network's ``state'' as the activity
pattern across its nodes, then every state is equally reachable by a network
with fully independent nodes. On the other hand, a network of identically
behaved nodes optimizes \textit{robustness}: any subset of nodes may be removed
from the network without any loss of function or expressive power, as long as
any single node remains. Presumably, most natural systems tend to occupy
positions between these extremes. We wondered: might the human brain
reconfigure itself to be more flexible or more robust according to ongoing
demands? In other words, might the brain reconfigure its connections or
behaviors under different circumstances to change its position along this
continuum?

Closely related to the above notions of expressiveness versus robustness are
measures of how much \textit{information} is contained in a given signal or
pattern, and how \textit{redundant} a signal is~\citep{Shan48}. Formally,
information is defined as the amount of uncertainty about a given variables'
outcomes (i.e., entropy), measured in \textit{bits}, or the optimal number of
yes/no questions needed to reduce uncertainty about the variable's outcomes to
zero. Highly complex systems with many degrees of freedom (i.e., high
flexibility and expressiveness), are more information-rich than simpler or more
constrained systems. The redundancy of a signal denotes the difference how
expressive the signal \textit{could} be (i.e., proportional to the number of
unique states or symbols used to transmit the signal) and the actual
information rate (i.e., the entropy of each individual state or symbol). If a
brain network's nodes are fully independent, then the number of bits required
to express a single activity pattern is proportional to the number of nodes.
The network would also be minimally redundant, since the status of every node
would be needed to fully express a single brain activity pattern. If a brain
network's nodes are fully coupled and identical, then the number of bits
required to express a single activity pattern is proportional to the number of
unique states or values any individual node can take on. Such a network would
be highly redundant, since knowing any individual node's state would be
sufficient to recover the full-brain activity pattern. Highly redundant systems
are also robust, since there is little information loss from losing any given
observation.

@@

  \caption{\textbf{Information content and compressibility.} \textbf{A. Variance
  explained for two images.} We applied principal components analysis to a
  photograph and drawing, treating each row of the images as ``observations.''
  Across columns, we identified the number of components required to explain
  100\%, 95\%, 75\%, or 50\% of the cumulative variance in each image (the 100\%
  columns denote the original images). The numbers of components are indicate in
  parentheses, and the resulting ``compressed'' images are displayed. \textbf{B.
  Representing two images with different numbers of components.} Using the same
  principal component decompositions as in Panel A, we computed the cumulative
  proportion of variance explained with 380 (original images), 20, 4, or 2
  components. \textbf{C. Cumulative variance explained versus number of
  components.} For the images displayed in Panels A and B, we plot the cumulative
  proportion of variance explained as a function of the number of components used
  to represent each image. \textbf{D. Information rate and compressibility.}
  Across multiple images, the information rate (i.e., the amount of information
  contained in each image; horizontal axis) is high when each individual pixel
  provides information that cannot be inferred from other pixels.
  High-information rate images tend to be high-resolution, and low-information
  rate images tend to be low-resolution. Compressibility is related to the
  difference between the information required to specify the original versus
  compressed images (vertical axis). Highly compressible images often contain
  predictable structure (redundancies) that can be leveraged to represent the
  images much more efficiently than in their original feature spaces.}

@@

The above studies are consistent with two potential descriptions of how
cognitive processes are reflects in brain activity patterns. One possibility is
that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns increases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings, or compressed representations of brain
recordings, should carry more information during complex or high-level (versus
simple or low-level) cognitive tasks.

@@

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. The horizontal red line denotes chance performance, and the
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows.
\textbf{C. Estimating inflection points.} We sought to identify an ``inflection
point'' for each decoding curve, denoting the number of components at which the
decoding curve asymptotes. We fit sigmoid functions to each decoding curve
(left sub-panel) and then computed the minimum number of components where the
second derivative of the sigmoid was both positive and less than a threshold
value of 0.0001. \textbf{D. Inflection points by condition.} Each dot displays
the number of components ($x$-axis) and decoding accuracy ($y$-axis) at one
condition's inflection point. All error ribbons denote bootstrap-estimated 95\%
confidence intervals.}

@@

To evaluate the relation between informativeness and compression for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data. Several patterns
were revealed by the analysis. First, in general (i.e., across experimental
conditions), decoding accuracy improves as the number of components increases.
However, decoding accuracy peaked at a higher level for experimental conditions
that exposed participants to cognitively richer stimuli. The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(XXX) =
XXX, p = XXX$; versus word: $t(XXX) = XXX, p = XXX$; versus rest: $t(XXX) =
XXX, p = XXX$), next highest for the ``paragraph'' condition (versus word:
$t(XXX) = XXX, p = XXX$; versus rest: $t(XXX) = XXX, p = XXX$), and next
highest for the ``word'' condition (versus rest: $t(XXX) = XXX, p = XXX$). 
This ordering implies that cognitively richer conditions evoke more stable brain
activity patterns across people.

@@

We also wondered how informativeness and compressibility in the different
experimental conditions might vary across brain networks. We used a network
parcellation identified by \cite{YeoEtal11} to segment the brain into seven
distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Fig.~\ref{fig:networks}A):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy (for the
intact condition) tended to increase
(Figs.~\ref{fig:networks}B,~\accuracyNetworks). This suggests that higher-order
networks may carry more content-relevant or stimulus-driven ``information.'' We
found no clear trends in the numbers of components at each network's inflection
point across networks or conditions
(Figs.~\ref{fig:networks}C,~\componentsNetworks). \textbf{JRM NOTE: expand on
this once we have the new figures.}

@@

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns are also more compressible, in that each individual
component provided more ``signal'' to temporal decoders relative to components
of data from less cognitively rich conditions (Fig.~\ref{fig:inflection}). Over
time (e.g., as the experiment progressed), these phenomena were strengthened.
Specifically, across story segments, data from more cognitively rich conditions
became more informative and compressible, and data from less cognitively rich
conditions became \textit{less} informative and compressible
(Fig.~\ref{fig:inflection-thirds}). We also repeated these analyses separately
for different brain networks. We found that networks traditionally associated
with higher-level cognitive functions tended to provide more informative brain
patterns than networks traditionally associated with lower level cognitive
functions (Fig.~\ref{fig:networks}). Finally, we examined the most dominant
components of the brain activity patterns from each experimental condition. We
used a reverse inference approach~\citep{RubiEtal17} to identify the terms in
the neuroimaging literature most commonly associated with the corresponding
maps. As summarized in Figure~\ref{fig:discussion}, we found that terms
associated with memory and sensory processing were associated with the
strongest components in all three story listening conditions. Terms associated
with sensory integration were associated with the strongest components in the
intact and paragraph-scrambled conditions. Terms associated with sentence
comprehension, emotion, and valence were associated with the strongest
components in the intact condition. Finally, terms associated with the default
mode network were associated with the strongest components in the
word-scrambled and resting state conditions. Taken together, our findings
indicate that the informativeness and compressibility of our brain activity
patterns are task-dependent, and these properties change systematically with
factors like cognitive richness and depth of processing.

@@

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10, SizeEtal18,
SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11, AdacEtal12,
BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04, DhamEtal08,
KorzEtal08, BrovEtal04}. Correlations or causal associations between different
brain regions simultaneously imply that full-brain activity patterns will be
compressible and also that those activity patterns will contain redundancies.
For example, the extent that activity patterns at one brain area can be
inferred or predicted from activity patterns at other
areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then the
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other region's activity patterns
that are predictive of the unreliable region.

Our findings that informativeness and compressibility change with task demands
may follow from task-dependent changes in full-brain correlation patterns. A
number of prior studies have found that so-called ``functional connectivity''
(i.e., correlational) patterns vary reliably across tasks, events, and
situations~\citep{SimoEtal16, ColeEtal14, SmitEtal09, OwenEtal21}. By examining
how these task-dependent changes in correlations affect informativeness and
compressibility, our work suggests a potential reason for why the statistical
structure of brain activity patterns might vary with cognitive task or with
cognitive demands. For lower-level tasks, or for tasks that require relatively
little ``deep'' cognitive processing, our brains may optimize activity patterns
for robustness and redundancy over expressiveness, for example to maximize
reliability. For higher-level tasks, or for tasks that require deeper cognitive
processing, our brains may sacrifice some redundancy in favor of greater
expressiveness.

In the information theory sense~\citep{Shan48}, when a signal is transmitted
using a fixed alphabet of ``symbols,'' the information rate decreases as the
signal is compressed (e.g., fewer symbols transmitted per unit time, using an
alphabet with fewer symbols, etc.). Our finding that each individual brain
component (symbol) becomes more informative as cognitive richness increases
suggests that the ``alphabet'' of brain activity patterns is also
task-dependent. Other work suggests that the representations that are
\textit{reflected} by brain activity patterns may also change with task
demands. For example, our brains may represent the same perceptual stimulus
differently depending on which aspects of the stimulus or which combinations of
features are relevant~\citep{MackEtal20}.

Different brain networks also varied in how informative and compressible their
activity patterns were across experimental conditions (e.g.,
Fig.~\ref{fig:networks}). This might reflect evolutionary optimizations that
reflect the relevant constraints or demands placed on those networks. One
possibility is that cortex is organized in a hierarchy of networks concerned
with different levels of processing or function. To the extent that different
levels of processing (e.g., low-level sensory processing versus ``deeper''
higher-level processing) reflect different stimulus
timescales~\citep[e.g.,][]{Mann20}, the network differences we observed might
also relate to the timescales ot which each network is maximally
sensitive~\citep{RegeEtal18, BaldEtal17,LernEtal11, HassEtal08}.

@@

Cognitive neuroscientists are still grappling with basic questions about the
fundamental ``rules'' describing how our brains respond, and about how brain
activity patterns and the associated cognitive representations and computations
are linked. We identified two aspects of brain activity patterns,
informativeness and compressibiilty, that appear to change systematically with
task demands and across brain networks. Our work helps to clarify how the
``neural code'' might be structured, and how the code might vary across tasks
and brain areas.

@@

We examined an fMRI dataset collected by \cite{SimoEtal16} that the authors
have made publicly available at
\href{http://arks.princeton.edu/ark:/88435/dsp015d86p269k}{arks.princeton.edu/ark:/88435/dsp015d86p269k}.
The dataset comprises neuroimaging data collected as participants listened to
an audio recording of a story (intact condition; 36 participants), listened to
temporally scrambled recordings of the same story (17 participants in the
paragraph-scrambled condition listened to the paragraphs in a randomized order
and 36 in the word-scrambled condition listened to the words in a randomized
order), or lay resting with their eyes open in the scanner (rest condition; 36
participants). Full neuroimaging details may be found in the original paper for
which the data were collected~\cite{SimoEtal16}. Procedures were approved by
the Princeton University Committee on Activities Involving Human Subjects, and
by the Western Institutional Review Board (Puyallup, WA). All subjects were
native English speakers with normal hearing and provided written informed
consent.

@@

Following our prior related work, we used HTFA~\cite{MannEtal18} to derive a
compact representation of the neuroimaging data. In brief, this approach
approximates the timeseries of voxel activations (44,415 voxels) using a much
smaller number of radial basis function (RBF) nodes~\citep[in this case, 700
nodes, as determined by an optimization procedure;][]{MannEtal18}). This
provides a convenient representation for examining full-brain activity patterns
and network dynamics. All of the analyses we carried out on the neuroimaging
dataset were performed in this lower-dimensional space. In other words, each
participant's data matrix, $\mathbf{X}$, was a number-of-timepoints ($T$) by 700
matrix of HTFA-derived factor weights (where the row and column labels were
matched across participants). Code for carrying out HTFA on fMRI data may be
found as part of the BrainIAK toolbox~\cite{CapoEtal17, KumaEtal21}, which may
be downloaded at \href{https://brainiak.org/}{brainiak.org}.

@@

We applied group PCA~\citep{SmitEtal14} separately to the HTFA-derived
representations of the data (i.e., factor loadings) from each experimental
condition. Specifically, for each condition, we considered the set of all
participants' $T$ by 700 factor weight matrices. We used group PCA to project
these 700-dimensional matrices into a series of $k$-dimensional spaces, for $k
\in \{3, 4, 5, ..., 700\}$. This yielded a set of number-of-participants
matrices, each with $T$ rows and $k$ columns.

@@

We sought to identify neural patterns that reflected participants' ongoing
cognitive processing of incoming stimulus information. As reviewed by
\cite{SimoEtal16}, one way of homing in on these stimulus-driven neural
patterns is to compare activity patterns across individuals. In particular,
neural patterns will be similar across individuals to the extent that the
neural patterns under consideration are stimulus-driven, and to the extent that
the corresponding cognitive representations are reflected in similar spatial
patterns across people~\cite{SimoChan20}. Following this logic, we used an
across-participant temporal decoding test developed by \cite{MannEtal18} to
assess the degree to which different neural patterns reflected ongoing
stimulus-driven cognitive processing across people. The approach entails using
a subset of the data to train a classifier to decode stimulus timepoints (i.e.,
moments in the story participants listened to) from neural patterns. We use
decoding (forward inference) accuracy on held-out data, from held-out
participants, as a proxy for the extent to which the inputted neural patterns
reflected stimulus-driven cognitive processing in a similar way across
individuals.

@@

To help interpret the brain activity patterns we found within the contexts of
other studies, we created summary maps of each principal component, for each
experimental condition, by summing together the 20 HTFA-derived RBF nodes (see
\textit{Hierarchical Topographic Factor Analysis}) with the highest absolute
value weights for each of the top 5 components (Figs.~\ref{fig:neurosynth-pca},
\synthThirds). We then carried out a meta analysis using
Neurosynth~\cite{RubiEtal17} to identify the 5 terms most commonly associated
with the given map.

@@

All of the code used to produce the figures and results in this manuscript,
along with links to the corresponding datasets, may be found at 
\href{https://github.com/ContextLab/pca_paper}{github.com/ContextLab/pca\_paper}.

@@

We acknowledge discussions with Rick Betzel, Emily Finn, and Jim Haxby. Our
work was supported in part by NSF CAREER Award Number 2145172 to J.R.M. The
content is solely the responsibility of the authors and does not necessarily
represent the official views of our supporting organizations. The funders had
no role in study design, data collection and analysis, decision to publish, or
preparation of the manuscript.

>>>>

\title{High-order cognition is supported by information-rich but compressible brain activity patterns}

@@

We applied dimensionality reduction algorithms and pattern classifiers to
functional neuroimaging data collected as participants listened to a story,
temporally scrambled versions of the story, or underwent a resting state
scanning session. These experimental conditions were intended to require
different depths of processing and inspire different levels of cognitive
engagement. We considered two primary aspects of the data. First, we treated
the maximum achievable decoding accuracy across participants as an indicator of
the ``informativeness'' of the recorded patterns. Second, we treated the number
of features (components) required to achieve a threshold decoding accuracy as a
proxy for the ``compressibility'' of the neural patterns (where fewer
components indicate greater compression). Overall, we found that the peak
decoding accuracy (achievable without restricting the numbers of features)
was highest in the intact (unscrambled) story listening condition.  However, the
number of features required to achieve comparable classification accuracy
was also lowest in the intact story listening condition.  Taken together, our work
suggests that our brain networks flexibly reconfigure according to ongoing task
demands, and that the activity patterns associated with higher-order cognition
and high engagement are both more informative and more compressible than the
activity patterns associated with lower-order tasks and lower levels of
engagement.

@@

Large-scale networks, including the human brain, may be conceptualized as
occupying one or more positions along on a continuum. At one extreme, every
node is fully independent from every other node. At the other extreme, all
nodes behave identically. Each extreme optimizes key properties of how the
network functions. When every node is independent, the network is maximally
\textit{expressive}: if we define the network's ``state'' as the activity
pattern across its nodes, then every state is equally reachable by a network
with fully independent nodes. On the other hand, a network of identically
behaved nodes optimizes \textit{robustness}: any subset of nodes may be removed
from the network without any loss of function or expressive power, as long as
any single node remains. Presumably, most natural systems tend to occupy
positions between these extremes. We wondered: might the human brain
reconfigure itself to be more flexible or more robust according to ongoing
demands? In other words, might the brain reconfigure its connections or
behaviors under different circumstances to change its position along this
continuum?

Closely related to the above notions of expressiveness versus robustness are
measures of how much \textit{information} is contained in a given signal or
pattern, and how \textit{redundant} a signal is~\citep{Shan48}. Formally,
information is defined as the amount of uncertainty about a given variables'
outcomes (i.e., entropy), measured in \textit{bits}, or the optimal number of
yes/no questions needed to reduce uncertainty about the variable's outcomes to
zero. Highly complex systems with many degrees of freedom (i.e., high
flexibility and expressiveness), are more information-rich than simpler or more
constrained systems. The redundancy of a signal denotes the difference between
how expressive the signal \textit{could} be (i.e., proportional to the number
of unique states or symbols used to transmit the signal) and the actual
information rate (i.e., the entropy of each individual state or symbol). If a
brain network's nodes are fully independent, then the number of bits required
to express a single activity pattern is proportional to the number of nodes.
The network would also be minimally redundant, since the status of every node
would be needed to fully express a single brain activity pattern. If a brain
network's nodes are fully coupled and identical, then the number of bits
required to express a single activity pattern is proportional to the number of
unique states or values any individual node can take on. Such a network would
be highly redundant, since knowing any individual node's state would be
sufficient to recover the full-brain activity pattern. Highly redundant systems
are also robust, since there is little total information loss due to removing
any given observation.

@@

  \caption{\textbf{Information content and compressibility.} \textbf{A.
  Variance explained for two images.} We applied principal components analysis
  to a photograph and drawing, treating the rows of the images as
  ``observations.'' Across columns, we identified the numbers of components
  required to explain 100\%, 95\%, 75\%, or 50\% of the cumulative variance in
  each image (the 100\% columns denote the original images). The numbers of
  components are indicate in parentheses, and the resulting ``compressed''
  images are displayed. \textbf{B. Representing two images with different
  numbers of components.} Using the same principal component decompositions as
  in Panel A, we computed the cumulative proportion of variance explained with
  380 (original images), 20, 4, or 2 components. \textbf{C. Cumulative variance
  explained versus number of components.} For the images displayed in Panels A
  and B, we plot the cumulative proportion of variance explained as a function
  of the number of components used to represent each image. \textbf{D.
  Information rate and compressibility.} Across multiple images, the
  information rate (i.e., the amount of information contained in each image;
  horizontal axis) is high when each individual pixel provides information that
  cannot be inferred from other pixels. High-information rate images tend to be
  high-resolution, and low-information rate images tend to be low-resolution.
  Compressibility is related to the difference between the information required
  to specify the original versus compressed images (vertical axis). Highly
  compressible images often contain predictable structure (redundancies) that
  can be leveraged to represent the images much more efficiently than in their
  original feature spaces.}

@@

The above studies are consistent with two potential descriptions of how
cognitive processes are reflected in brain activity patterns. One possibility
is that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns increases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings, or compressed representations of brain
recordings, should carry more information during complex or high-level (versus
simple or low-level) cognitive tasks.

@@

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. The horizontal red line denotes chance performance, and the
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows.
\textbf{C. Estimating inflection points.} We sought to identify an ``inflection
point'' for each decoding curve, denoting the number of components at which the
decoding curve asymptotes. We fit sigmoid functions to each decoding curve
(left sub-panel) and then computed the minimum number of components for which
the second derivative of the sigmoid was both positive and less than a
threshold value of 0.0001. \textbf{D. Inflection points by condition.} Each dot
displays the number of components ($x$-axis) and decoding accuracy ($y$-axis)
at one condition's inflection point. All error ribbons denote
bootstrap-estimated 95\% confidence intervals.}

@@

To evaluate the relation between informativeness and compressibiilty for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data. Several patterns
were revealed by the analysis. First, in general (i.e., across experimental
conditions), decoding accuracy improves as the number of components increases.
However, decoding accuracy peaked at higher levels for experimental conditions
that exposed participants to cognitively richer stimuli. The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(XXX) =
XXX, p = XXX$; versus word: $t(XXX) = XXX, p = XXX$; versus rest: $t(XXX) =
XXX, p = XXX$), next highest for the ``paragraph'' condition (versus word:
$t(XXX) = XXX, p = XXX$; versus rest: $t(XXX) = XXX, p = XXX$), and next
highest for the ``word'' condition (versus rest: $t(XXX) = XXX, p = XXX$). This
ordering implies that cognitively richer conditions evoke more stable brain
activity patterns across people.

@@

We also wondered how informativeness and compressibility in the different
experimental conditions might vary across brain networks. We used a network
parcellation identified by \cite{YeoEtal11} to segment the brain into seven
distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Fig.~\ref{fig:networks}A):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy (for the
intact condition) tended to increase (Fig.~\ref{fig:networks}B). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the numbers of components at each
network's inflection point across networks or conditions
(Fig.~\ref{fig:networks}C).

@@

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible (Fig.~\ref{fig:inflection-thirds}).
We also repeated these analyses separately for different brain networks. We
found that networks traditionally associated with higher-level cognitive
functions tended to provide more informative brain patterns than networks
traditionally associated with lower level cognitive functions
(Fig.~\ref{fig:networks}). Finally, we examined the most dominant components of
the brain activity patterns from each experimental condition. We used a reverse
inference approach~\citep{RubiEtal17} to identify the terms in the neuroimaging
literature most commonly associated with the corresponding maps. As summarized
in Figure~\ref{fig:discussion}, we found that terms associated with memory and
sensory processing were associated with the strongest components in all three
story listening conditions. Terms associated with sensory integration were
associated with the strongest components in the intact and paragraph-scrambled
conditions. Terms associated with sentence comprehension, emotion, and valence
were associated with the strongest components in the intact condition. Finally,
terms associated with the default mode network were associated with the
strongest components in the word-scrambled and resting state conditions. Taken
together, our findings indicate that the informativeness and compressibility of
our brain activity patterns are task-dependent, and these properties change
systematically with factors like cognitive richness and depth of processing.

@@

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10, SizeEtal18,
SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11, AdacEtal12,
BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04, DhamEtal08,
KorzEtal08, BrovEtal04}. Correlations or causal associations between different
brain regions simultaneously imply that full-brain activity patterns will be
compressible and also that those activity patterns will contain redundancies.
For example, the extent ot which activity patterns at one brain area can be
inferred or predicted from activity patterns at other
areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then the
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region.

Our findings that informativeness and compressibility change with task demands
may follow from task-dependent changes in full-brain correlation patterns. A
number of prior studies have found that so-called ``functional connectivity''
(i.e., correlational) patterns vary reliably across tasks, events, and
situations~\citep{SimoEtal16, ColeEtal14, SmitEtal09, OwenEtal21}. By examining
how these task-dependent changes in correlations affect informativeness and
compressibility, our work suggests a potential reason why the statistical
structure of brain activity patterns might vary with cognitive task or with
cognitive demands. For lower-level tasks, or for tasks that require relatively
little ``deep'' cognitive processing, our brains may optimize activity patterns
for robustness and redundancy over expressiveness, for example to maximize
reliability. For higher-level tasks, or for tasks that require deeper cognitive
processing, our brains may sacrifice some redundancy in favor of greater
expressiveness.

In the information theory sense~\citep{Shan48}, when a signal is transmitted
using a fixed alphabet of ``symbols,'' the information rate decreases as the
signal is compressed (e.g., fewer symbols transmitted per unit time, using an
alphabet with fewer symbols, etc.). Our finding that each individual brain
component (symbol) becomes more informative as cognitive richness increases
suggests that the ``alphabet'' of brain activity patterns is also
task-dependent. Other work suggests that the representations that are
\textit{reflected} by brain activity patterns may also change with task
demands. For example, our brains may represent the same perceptual stimulus
differently depending on which aspects of the stimulus or which combinations of
features are task-relevant~\citep{MackEtal20}.

Different brain networks also varied in how informative and compressible their
activity patterns were across experimental conditions (e.g.,
Fig.~\ref{fig:networks}). This might follow from evolutionary optimizations
that reflect the relevant constraints or demands placed on those networks. One
possibility is that cortex is organized in a hierarchy of networks ``concerned
with'' or selective to different levels of processing or function. To the
extent that different levels of processing (e.g., low-level sensory processing
versus ``deeper'' higher-level processing) reflect different stimulus
timescales~\citep[e.g.,][]{Mann20}, the network differences we observed might
also relate to the timescales at which each network is maximally
sensitive~\citep{RegeEtal18, BaldEtal17,LernEtal11, HassEtal08}.

@@

Cognitive neuroscientists are still grappling with basic questions about the
fundamental ``rules'' describing how our brains respond, and about how brain
activity patterns and the associated underlying cognitive representations and
computations are linked. We identified two aspects of brain activity patterns,
informativeness and compressibiilty, that appear to change systematically with
task demands and across brain networks. Our work helps to clarify how the
``neural code'' might be structured, and how the code might vary across tasks
and brain areas.

@@

We examined an fMRI dataset collected by \cite{SimoEtal16} that the authors
have made publicly available at
\href{http://arks.princeton.edu/ark:/88435/dsp015d86p269k}{arks.princeton.edu/ark:/88435/dsp015d86p269k}.
The dataset comprises neuroimaging data collected as participants listened to
an audio recording of a story (intact condition; 36 participants), listened to
temporally scrambled recordings of the same story (17 participants in the
paragraph-scrambled condition listened to the paragraphs in a randomized order
and 36 in the word-scrambled condition listened to the words in a randomized
order), or lay resting with their eyes open in the scanner (rest condition; 36
participants). Full neuroimaging details may be found in the original paper for
which the data were collected~\citep{SimoEtal16}. Procedures were approved by
the Princeton University Committee on Activities Involving Human Subjects, and
by the Western Institutional Review Board (Puyallup, WA). All subjects were
native English speakers with normal hearing and provided written informed
consent.

@@

Following our prior related work, we used HTFA~\citep{MannEtal18} to derive a
compact representation of the neuroimaging data. In brief, this approach
approximates the timeseries of voxel activations (44,415 voxels) using a much
smaller number of radial basis function (RBF) nodes~\citep[in this case, 700
nodes, as determined by an optimization procedure;][]{MannEtal18}). This
provides a convenient representation for examining full-brain activity patterns
and network dynamics. All of the analyses we carried out on the neuroimaging
dataset were performed in this lower-dimensional space. In other words, each
participant's data matrix was a number-of-timepoints ($T$) by 700 matrix of
HTFA-derived factor weights (where the row and column labels were matched
across participants). Code for carrying out HTFA on fMRI data may be found as
part of the BrainIAK toolbox~\citep{CapoEtal17, KumaEtal21}, which may be
downloaded at \href{https://brainiak.org/}{brainiak.org}.

@@

We applied group PCA~\citep{SmitEtal14} separately to the HTFA-derived
representations of the data (i.e., factor loadings) from each experimental
condition. Specifically, for each condition, we considered the set of all
participants' $T$ by 700 factor weight matrices. We used group PCA to project
these 700-dimensional matrices into a series of shared $k$-dimensional spaces,
for $k \in \{3, 4, 5, ..., 700\}$. This yielded a set of number-of-participants
matrices, each with $T$ rows and $k$ columns.

@@

We sought to identify neural patterns that reflected participants' ongoing
cognitive processing of incoming stimulus information. As reviewed by
\cite{SimoEtal16}, one way of homing in on these stimulus-driven neural
patterns is to compare activity patterns across individuals. In particular,
neural patterns will be similar across individuals to the extent that the
neural patterns under consideration are stimulus-driven, and to the extent that
the corresponding cognitive representations are reflected in similar spatial
patterns across people~\citep{SimoChan20}. Following this logic, we used an
across-participant temporal decoding test developed by \cite{MannEtal18} to
assess the degree to which different neural patterns reflected ongoing
stimulus-driven cognitive processing across people. The approach entails using
a subset of the data to train a classifier to decode stimulus timepoints (i.e.,
moments in the story participants listened to) from neural patterns. We use
decoding (forward inference) accuracy on held-out data, from held-out
participants, as a proxy for the extent to which the inputted neural patterns
reflected stimulus-driven cognitive processing in a similar way across
individuals.

@@

To help interpret the brain activity patterns we found within the contexts of
other studies, we created summary maps of each principal component, for each
experimental condition, by summing together the 20 HTFA-derived RBF nodes (see
\textit{Hierarchical Topographic Factor Analysis}) with the highest absolute
value weights for each of the top 5 components (Figs.~\ref{fig:neurosynth-pca},
\synthThirds). We then carried out a meta analysis using
Neurosynth~\citep{RubiEtal17} to identify the 5 terms most commonly associated
with the given map.

@@

All of the code used to produce the figures and results in this manuscript,
along with links to the corresponding data, may be found at
\href{https://github.com/ContextLab/pca_paper}{github.com/ContextLab/pca\_paper}.

@@

We acknowledge discussions with Rick Betzel, Luke Chang, Emily Finn, and Jim
Haxby. Our work was supported in part by NSF CAREER Award Number 2145172 to
J.R.M. The content is solely the responsibility of the authors and does not
necessarily represent the official views of our supporting organizations. The
funders had no role in study design, data collection and analysis, decision to
publish, or preparation of the manuscript.
